Modular analysis of gene expression data with R

### 2010-03-15 isanote/answers_to_reviewers.tex

We have added these references.

\begin{quote}
2.      The sample score plots example given in the EISA tutorial
(\url{http://www2.unil.ch/cbg/index.php?title=EISA\_tutorial}, section 6.5)
is not sufficiently clear. What are the scores represented by the
horizontal lines and how can one calculate them? Moreover, how can one
separate B-cell and T-cell leukemia samples using this plot? 
\end{quote}

We have also explained sample score plots better in the tutorial. 
Please note, that the tutorials do not aim to explain all details 
about the ISA functions; the reference manual, that is included in the
package, is the place for that.

The module plotted in section 6.5 indeed cannot be used to separate
B-cell and T-cell samples, we point this out in the text.

@@

1.      The authors describe methods that perform enrichment tests for
the gene sets corresponding to the ISA modules against various
databases. In particular, they perform enrichment analysis against GO
database, KEGG pathway database, chromosomes and the TargetScan \citep{targetscan}
database. One may also want to investigate the correspondence of
biclusters and protein-protein interaction networks (which have been
derived from other types of data than gene expression data), e.g. the
DIP database \citep{salwinski04}. An example for such analysis can be
found in \citet{prelic06}.
\end{quote}

We have added new classes and functions to perform enrichment analysis
on the ISA modules, against any gene label or database that the user
provides. These functions are explained in the manual page of the
\Rfunction{ISAEnrichment} function.

\begin{quote}
2.      The heatmap image of a transcription module obtained by any
one of the two methods presented in the tutorial should include a
color legend. 
\end{quote}

We have added the color legend to the first heatmap. We removed the 
second heatmap from the tutorial entirely, as adding a heatmap to 
it would be too cumbersome for the user.

### 2010-03-15 isanote/modular.tex
ISA paper: added Sven's changes.

The output of ISA is a collection of potentially overlapping
modules. Every module contains genes, that are up- and/or
down-regulated, in samples that belong to the module.
In every module, each gene and each sample is attributed a
score between minus one and one, that reflects the strength of the
association with the module. Moreover, if the scores of two genes of a
module have the same sign, then they are correlated (across the
samples of the module), opposite signs mean
anti-correlation. Similarly, if two sample
scores have the same sign, then these samples are correlated (across
the genes of the module), opposite signs indicate anti-correlation.

For other biclustering algorithms, please see
\citet{cheng00,getz00,califano00,sharan02,tanay04,barkow06}.

@@

\begin{figure*}
\centering
\includegraphics[width=0.842\textwidth]{isa2workflow3}
\caption{\subfig{(A)} Work flow of a typical modular analysis with the
  \Rpackage{eisa} package. See text for details.
  Subfigures \subfig{B} and \subfig{C} were generated
  using the acute lymphoblastic leukemia data set, see
  \citep{chiaretti04} and the \Rpackage{ALL} R package.
  \subfig{(B)} Heatmap for a single module, showing coherent
  expression of the genes across the samples. The red lines are the gene and
  sample scores.
  \subfig{(C)} Module tree. Each rectangle is a module; see the
  definition of the edges in the text. Modules are colored
  according to their Gene Ontology enrichment $p$-values, the codes of
  the enriched GO categories are shown in the top-left corner of the
  rectangles. The top-right corner shows the number of genes and
  conditions in the module. The gene thresholds used for finding
  the modules is shown on the horizontal axes.
}
\label{fig:workflow}
\end{figure*}

@@

% Batch correction
\emph{Batch correction.}
To study the global organization of a transcription program including
many aspects of transcriptional regulation one often combines several
microarray experiments into a single dataset. In such a case
additional data normalization is crucial to reduce the bias due to
constituent datasets. Several methods address this challenge,
see e.g.~\citep{johnson07} for an algorithm that has a GNU R
implementation.

% Filtering
\emph{Gene filtering.}
Genes that have very low expression levels in all samples, carry little if
any information and may reflect ineffective array probes, etc. Since these
genes are likely to contribute only noise to the
analysis~\citep{hackstadt09}, we suggest removing them before running
the module identification of the ISA.

% Normalization.
\emph{ISA normalization.} (Step 1 on Fig.~\ref{fig:workflow}.)
In each iteration the ISA computes thresholded weighted sums of expression
levels over either genes or samples. Since different genes 
typically show different levels of base
expression and variance, it is important to standardize expression
levels to $Z$-scores. The ISA uses two sets of $Z$-scores, one
calculated for each gene across all samples, the other for each sample
across all genes.

>>>>

The output of ISA is a collection of potentially overlapping
modules. Every module contains genes, that are over- and/or
under-expressed, in samples that belong to the module.
In every module, each gene and each sample is attributed a
score between minus one and one, that reflects the strength of the
association with the module. Moreover, if the scores of two genes of a
module have the same sign, then they are correlated (across the
samples of the module), opposite signs mean
anti-correlation. Similarly, if two sample
scores have the same sign, then these samples are correlated (across
the genes of the module), opposite signs indicate anti-correlation.

For other biclustering algorithms, please see e.g.
\citet{cheng00,getz00,califano00,sharan02,tanay04,barkow06} and
\citet{ihmels04} for a review.

@@

\begin{figure*}
\centering
\includegraphics[width=0.842\textwidth]{isa2workflow3}
\caption{\subfig{(A)} Work flow of a typical modular analysis with the
  \Rpackage{eisa} package. See text for details.
  Subfigures \subfig{B} and \subfig{C} were generated
  using the acute lymphoblastic leukemia data set, see
  \citep{chiaretti04} and the \Rpackage{ALL} R package.
  \subfig{(B)} Heatmap for a single module, showing coherent
  expression of the genes across the samples. The red lines are the gene and
  sample scores.
  \subfig{(C)} Module tree. Each rectangle is a module; its numeric id
  is shown on top of it. See the definition of the edges in the
  text. Modules are colored 
  according to their Gene Ontology enrichment $p$-values, the codes of
  the enriched GO categories are shown in the top-left corner of the
  rectangles. The top-right corner shows the number of genes and
  conditions in the module. The gene thresholds used for finding
  the modules is shown on the horizontal axes.
}
\label{fig:workflow}
\end{figure*}

@@

% Batch correction
\emph{Batch correction.}
To study the global organization of a transcription program including
many aspects of transcriptional regulation one often combines several
microarray experiments into a single dataset. In such a case
additional data normalization is crucial to reduce the bias due to
the constituent datasets. Several methods address this challenge,
see e.g.~\citep{johnson07} for an algorithm that has a GNU R
implementation.

% Filtering
\emph{Gene filtering.}
Genes that have very low expression levels in all samples, carry little if
any information and may reflect ineffective array probes, etc. Since these
genes are likely to contribute mostly noise to the
analysis~\citep{hackstadt09}, we suggest removing them before running
the module identification of the ISA.

% Normalization.
\emph{ISA normalization.} (Step 1 in Fig.~\ref{fig:workflow}.)
In each iteration the ISA computes thresholded weighted sums of expression
levels over either genes or samples. Since different genes 
typically show different levels of base
expression and variance, it is important to standardize expression
levels to $Z$-scores. The ISA uses two sets of $Z$-scores, one
calculated for each gene across all samples, the other for each sample
across all genes.
